# Acknowledgements

*"An alle erschienenen und unerschienenen Erscheinungen
\newline
Menschen, Tiere, Meinungen, Gefühle, Fraktale
\newline
Zikaden, Mutanten, Schamanen, Fackelträger
\newline
Geistreisende und mikroenzyklopen-jagende Weichorganismen
\newline
Es ist an der Zeit einen Kreis zu bilden"*^[To all manifested and unmanifested manifestations, humans, animals, opinions, feelings, fractals, cicadas, mutants, shamans, torchbearers, astral-travellers and microencyclope-hunting molluscs, the time has come to form a circle.]
\newline
	                               - Käptn Peng & Die Tentakel von Delphi, Der Anfang ist nah (The beginning is nigh)

And what a journey it has been, this Phd'fying for the past five plus years. Sitting in front of the screen right now, I never really realised how much I've relied on all the amazing people and things around me to make it this far. Let's begin, starting from the office outwards in an ever growing radius from where I sit right now.

To my office-mates: Daniel who taught me the value of a nice silent office and the joy of handling bats, Theresa - your willingness to help and chat about all things science and random, (and for introducing me to Käptn Peng). Danke Danke. To all other former and current AFEG and associated members down the corridor (Aiqing, Arun,both Claires, Dylan, Erin, Eryn, Klemen, Laura, Lena, Leo, Neetash, Stefan, Toni, Verena) thank you all for forming a great and supportive and enthu environment to come to work every day. A big shout-out to Toni for the random conversation that made the whole 2018 field season possible at all! Holger, it has been awesome working with you - and I owe you a lot for teaching me the value of good presentation and rigour in science!

Haus 11, the out-house where all the cool people are, here's to all of you current and former Haus 11'ers for the random corridor conversations and barbecue evenings! Without the Narada/Hermes/Loki messenger-like figures (who roam across the three worlds and are bearers of all kinds of knowledge and bodings) Nicole Drenkard and Diana Werner, many things would not have been possible logistically and administratively - here's a big thanks to both of you. Joerg at the Pfore : quatsch'ing with you has always been a joy every time I came to pick up chocolate or a parcel. The Max-Planck Institute for Ornithology, my place of work, I thank for having the best work-place conditions (and location) and support systems I have ever experienced. I always looked forward to arriving at work. From the people around the instit now to all things and beings around: the peaceful snow outside my window, the lake, the beautiful cycle-path to the institute, and the occasional deer eating away peacefully - I am in some way thankful to these experiences and many more that have directly or indirectly come to this point in time and space.

To my more distantly located colleagues in my graduate school, the IMPRS for Organismal Biology and all of you at the Ghetto - I thank you all for forming a supportive net to share the joy and vent the frustrations of a life in science. Special thanks to the grad school coordinators Maeggi Hieber-Ruiz, Fransisca Rosa Mende and Corinna Loes for readily providing support whenever asked for. Reni, for providing an oasis to retreat from work while Qwirkl-ing away, and Dieter for being the amazing conversationalist and story-teller he is. 

My extended family in Germany: Apurva-Sindhu, Deepu-Ravi. My gratitude to you for creating a home away from home and a second base-camp over the past many years with their hospitality, food and love! Parts of this thesis literally wouldn't be written without all encouragement that came from giving me the space and time to write away in your homes along with all spiritual nourishment from the hot cups of chai and bonda-bajji's that came with it.

Neetash, for simultaneously playing the multiple roles of colleague, friend and wife over the past few years - I can not say enough, and perhaps words will not do justice! I can only be thankful that you have been by my side through the ups and downs as I tried to phd my way through to this point. Amma-Appa, for constantly supporting all my weird ideas of becoming, at various points, a vet and/or a 'zoologist' (even before I actually knew what the word meant), etc! Praj - for being my go-to person in times of joy and confusion for work and life related matters. I am here only because of the unquestioning 'I-have-your-back' attitude you all have shown at each point of confusion, frustration and joy in this journey till now. 

And finally, I would not be here without all the bats flying through the night and in the caves far far away in Bulgaria. All the bats that hung around as I fumbled with cables with numb hands on cold autumn nights, as I learnt to patiently deal with (hundreds of?) metres of tangled cables every night, and set up equipment hoping they'd land up. When they did it was always a pleasure to watch and hear them. So here's to all you hovercraft Daubis, hyper pipistrelles and gliding Mouse-eared bats, none of this would have been possible without all of you - and this is looking to spend more years to come understanding you weird creatures! 

```{r, include=FALSE}
# \foreignlanguage{bulgarian}{Благодаря},  \foreignlanguage{kannada}{ಧನ್ಯವಾದಗಳು}.
```

To everyone and everything mentioned above, and especially to all those who I have not specifically mentioned, including you my reader - I convey my gratitude: thank you, Danke, <TOCOMPLETE><TOCOMPLETE>


```{r, include=FALSE}
# inti nimma, Thejasvi 
# January 2021
```



